An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2021-10-28 ju-hofmann/ju-hofmann.Rmd
final proofread?

<!-- According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 73 percent (73.7) of the homicides for which the FBI received weapons data in 2019 involved the use of firearms" [@fbiucr], and firearm examination could play an important role in all these cases.  -->
<!-- \hh{XXX the new data for 2020 came out on Sep 27 - homicides are up by close to 30%. We might want to mention that.} -->
<!-- https://s3-us-gov-west-1.amazonaws.com/cg-d4b776d0-d898-4153-90c8-8336f86bdfec/CIUS/downloads/2020/expanded-homicide-2020.zip -->
<!-- Firearm examination has  played an important role in  convictions [@fbiucr] in the United States criminal justice system. \hh{Do we have a number or a percentage of cases where firearm examination has played a role? or maybe gun violence? XXX} --> <!-- https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/expanded-homicide --> 
\wj{According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 76 percent (76.7) of the homicides for which the FBI received weapons data in 2020 involved the use of firearms", and the number of cases that involved the use of firearms in 2020 increased by close to 30 percent (29.3) compared to that in 2019 [@fbiucr]. Firearm examination could play an important role in all these cases. }
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

<!-- \hh{XXX could you summarize the main contribution of each of these papers in at most a sentence?} -->
A report published by the @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field.
\wj{The database} of @nistdb and data sets of @brundage, @hamby, @Hamby:2019 provide great open data resources for research. Algorithms that can be validated and tested and generate quantitative results are developed for 2D and 3D surface texture [@song2005], tool marks [@ChumbleyL_Scott2010VoTM], and striae on Land Engraved Areas (LEAs) of bullets [@aoas, @pmid30444940, @cmps]. 
Note that this is not an exhaustive list of these efforts.
\wj{The notion of bullet signatures is one of the results of these efforts, and we will use it to explain the CMPS algorithm and showcase} the \pkg{cmpsR} implementation.
Bullet signatures can be extracted from bullet land engraved areas (LEAs), typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Since bullet signatures can numerically describe the striation marks on the bullet and can be understood by algorithms, they played an important role in bringing objectivity into the field.
In the following sections we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

Using the sum of squares ratio $V$ as an evaluation metric, we are able to construct a pipeline that aims to find the optimal parameter values for the CMPS algorithm by maximizing the sum of squares ratio.
\wj{Note that other measures of an algorithm, such as the accuracy and the AUC (Area Under the Curve), are also important and useful. But when algorithms achieve 100 percent accuracy and AUC of value 1, we need another measure like the sum of squares ratio to further distinguish the performance of algorithms.} In the following section, we will use the sum of squares ratio to compare the CMPS metrics introduced earlier and investigate effects of different parameter settings.

@@

\hh{XXX Additional line for the conclusion: Hamby sets 44 and 252 were taken with different resolution scanning devices. The bullets for Hamby 44 and Hamby 252 are fired from the same ten consecutively rifled P-85 Ruger barrels. The difference in parameter settings is therefore quite surprising. As next steps we need validation studies - trying out the CMPS algorithm on firearms and ammunition combinations that are different to test the limits of the algorithm. XXX shoutout to Susan's paper? - will look up the reference }

@@

\wj{Additionally, we propose new metrics based on the CMPS scores and compare the new metrics with the existing metrics. 
We also introduce a principled evaluation framework of algorithmic results using  a measure based on the sum of squares ratio. 
We showcase the implementation with two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare the CMPS algorithm using different sets of parameters and different metrics, following the evaluation framework based on the sum of squares ratio. }
<!-- Additionally, we apply the \pkg{cmpsR} package to two datasets in Hamby study (Hamby set 252 and Hamby set 44).  -->
The results obtained are promising:  we were both able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known match (KM) comparisons and known non-match (KNM) comparisons in another bullet study.
\wj{However, the comparisons among different sets of parameters suggest that the optimal choice for parameter settings vary between different datasets.}

\wj{Note that Hamby sets 44 and 252 were taken with different resolution scanning devices. The bullets for Hamby 44 and Hamby 252 are fired from the same ten consecutively rifled P-85 Ruger barrels. The difference in parameter settings is therefore quite surprising. As next steps we need validation studies - trying out the CMPS algorithm on firearms and ammunition combinations that are different to test the limits of the algorithm. The evaluation framework we proposed based on the sum of squares ratio will facilitate such studies and other validation of the algorithm.}

\wj{Comparisons of the R implementation of the CMPS algorithm with the random forest model proposed by @aoas  suggest that adding the CMPS score as an additional feature in the random forest model might add further separation between known match (KM) comparisons and known non-match (KNM) comparisons.}

\wj{The open-source implementation of the CMPS algorithm provided by the \pkg{cmpsR} package is just one step towards in the framework of open science. Open science is particularly important to fields like forensic science where transparency and accuracy are critical to fair justice. 
Open-source implementations not only allow a peer-review but also facilitate further research, such as parameter cross-validations, method comparisons, and the development of statistical methods for modeling KM and KNM CMPS score distributions, which can then be used for error-rate estimations and fair applications that were called for by the PCAST [@pcast] report.
However, having open-source implementations is not enough for open science. 
In order to actually build the world of open science, many other efforts, such as open evaluation, open access publication, open educational resources, etc., are still needed. 
The database built and maintained by the National Institute of Standards and Technology is a good example of open data. 
What we are aiming for is an open system that is able to collect open results, compare multiple algorithms using multiple datasets, and evaluate algorithmic variation and accuracy. 
The evaluation metric we proposed in this paper can be used to compare different algorithms or even different datasets and is a snippet of this open system. 
But the core of this open system is the open science culture and contributions of the community.}

>>>>

<!-- According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 73 percent (73.7) of the homicides for which the FBI received weapons data in 2019 involved the use of firearms" [@fbiucr], and firearm examination could play an important role in all these cases.  -->
<!-- \hh{XXX the new data for 2020 came out on Sep 27 - homicides are up by close to 30%. We might want to mention that.} -->
<!-- https://s3-us-gov-west-1.amazonaws.com/cg-d4b776d0-d898-4153-90c8-8336f86bdfec/CIUS/downloads/2020/expanded-homicide-2020.zip -->
<!-- Firearm examination has  played an important role in  convictions [@fbiucr] in the United States criminal justice system. \hh{Do we have a number or a percentage of cases where firearm examination has played a role? or maybe gun violence? XXX} --> <!-- https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/expanded-homicide --> 
According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 76 percent (76.7) of the homicides for which the FBI received weapons data in 2020 \hh{is this really 2020 or should it be 2019?}involved the use of firearms", and the number of cases that involved the use of firearms in 2020 increased by close to 30 percent (29.3) compared to that in 2019 [@fbiucr]. Firearm examination could play an important role in all these cases. 
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

<!-- \hh{XXX could you summarize the main contribution of each of these papers in at most a sentence?} -->
A report published by the @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field.
The database of @nistdb with digital 3d topographic scans from various studies by @brundage, @hamby, @Hamby:2019 and others, provides a great resource for researchers. Algorithms that can be validated and tested and generate quantitative results are developed for 2D and 3D surface texture [@song2005], tool marks [@ChumbleyL_Scott2010VoTM], and striae on Land Engraved Areas (LEAs) of bullets [@aoas, @pmid30444940, @cmps]. 
<!--Note that this is not an exhaustive list of these efforts.-->
The notion of bullet signatures is one of the results of these efforts, and we will use it to explain the CMPS algorithm and showcase the \pkg{cmpsR} implementation.
Bullet signatures can be extracted from bullet land engraved areas (LEAs), typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Bullet signatures capture striation marks on the bullet in a numeric format and therefore serve as the foundation for algorithms. These scans played an important role in bringing objectivity into the field.
In the following sections we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

Using the sum of squares ratio $V$ as an evaluation metric, we are able to construct a pipeline that aims to find the optimal parameter values for the CMPS algorithm by maximizing the sum of squares ratio.
Note that other measures of an algorithm, such as the accuracy and the AUC (Area Under the Curve), are also important and useful. But when algorithms achieve 100 percent accuracy and AUC value of 1, we need other measures such as the sum of squares ratio to further distinguish the performance of algorithms. In the following section, we will use the sum of squares ratio to compare the CMPS metrics introduced earlier and investigate effects of different parameter settings.

@@

@@

Additionally, we propose two new metrics based on the CMPS scores and compare the new metrics with the existing metrics. 
We also introduce a principled evaluation framework of algorithmic results using  a measure based on the sum of squares ratio. 
We showcase the implementation with two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare the CMPS algorithm using different sets of parameters and different metrics, following the evaluation framework based on the sum of squares ratio. 
<!-- Additionally, we apply the \pkg{cmpsR} package to two datasets in Hamby study (Hamby set 252 and Hamby set 44).  -->
The results obtained are promising:  we were both able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known match (KM) comparisons and known non-match (KNM) comparisons in another bullet study.
However, the comparisons among different sets of parameters suggest that the optimal choice for parameter settings vary between different datasets.

Note, that the main difference between Hamby sets 44 and 252 is that they were taken with different resolution scanning devices. The bullets for Hamby 44 and Hamby 252 are fired from the same ten consecutively rifled P-85 Ruger barrels. The difference in parameter settings for optimal differentiation between same-source comparisons and different-source comparisons is therefore quite surprising. As next steps we need validation studies, similar to @vanderplas - trying out the CMPS algorithm on firearms and ammunition combinations that are different to test the limits of the algorithm. The evaluation framework we proposed based on the sum of squares ratio will facilitate such studies and other validation of the algorithm.

Comparisons of the R implementation of the CMPS algorithm with the random forest model proposed by @aoas  suggest that adding the CMPS score as an additional feature in the random forest model might add further separation between known match (KM) comparisons and known non-match (KNM) comparisons.

The open-source implementation of the CMPS algorithm provided by the \pkg{cmpsR} package is just one step towards in the framework of open science. Open science is particularly important to fields like forensic science where transparency and accuracy are critical to fair justice. 
Open-source implementations not only allow a peer-review but also facilitate further research, such as parameter cross-validations, method comparisons, and the development of statistical methods for modeling KM and KNM CMPS score distributions, which can then be used for error-rate estimations and fair applications that were called for by the PCAST [@pcast] report.
However, having open-source implementations is not enough for open science. 
In order to actually build the world of open science, many other efforts, such as open evaluation, open access publication, open educational resources, etc., are still needed. 
The database built and maintained by the National Institute of Standards and Technology is a good example of open data. 
What we are aiming for is an open system that is able to collect open results, compare multiple algorithms using multiple datasets, and evaluate algorithmic variation and accuracy. 
The evaluation metric we proposed in this paper can be used to compare different algorithms or even different datasets and is a snippet of this open system. 
But the core of this open system is the open science culture and contributions of the community.

### 2021-10-28 ju-hofmann/ju-hofmann.tex
final proofread?

According to the Uniform Crime Reporting Program of FBI (Federal Bureau
of Investigation), ``More than 73 percent (73.7) of the homicides for
which the FBI received weapons data in 2019 involved the use of
firearms'' \citep{fbiucr}, and firearm examination could play an
important role in all these cases. \textbackslash hh\{XXX the new data
for 2020 came out on Sep 27 - homicides are up by close to 30\%. We
might want to mention that.\} An important task of firearm examination
is to answer the question of whether two pieces of evidence come from
the same source or whether a piece of evidence matches a sample obtained
from a specific firearm. Here, in particular, we are interested to
determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of
striation marks acquired during the firing process as bullets are
propelled through the barrel. Current state of art sees firearms
examiners make an assessment of similarity based on a visual comparison,
generally, using a comparison microscope \citep{afte}. This practice has
been criticized for its lack of objectivity and the associated problem
of determining valid error rates \citep{pcast}.

A report published by the \citet{nrc} states that ``{[}m{]}uch forensic
evidence-including, for example, bite marks and firearm and toolmark
identification---is introduced in criminal trials without any meaningful
scientific validation, determination of error rates, or reliability
testing to explain the limits of the discipline.'' To overcome those
criticisms and concerns, researchers have been making an effort to build
databases and develop frameworks and algorithms that bring an objective
and quantitative assessment into the field
\hh{XXX could you summarize the main contribution of each of these papers in at most a sentence?}
\citetext{\citealp[\citet{brundage}, \citet{hamby}, \citet{Hamby:2019},
\citet{song2005}, \citet{ChumbleyL_Scott2010VoTM}, \citet{aoas},
\citet{pmid30444940}]{nistdb}; \citealp{cmps}}. Note that this is not an
exhaustive list of these efforts. The notion of bullet signatures is one
of the results of these efforts. Bullet signatures can be extracted from
bullet land engraved areas (LEAs), typically one for each land engraved
area, and how those bullet signatures are extracted from scans will be
discussed in the Background section. Since bullet signatures can
numerically describe the striation marks on the bullet and can be
understood by algorithms, they played an important role in bringing
objectivity into the field. In the following sections we will: review
the background of bullet signature comparisons, discuss how we followed
the idea described by \citet{cmps} for the implementation of the CMPS
algorithm, propose new metrics based on the CMPS score and a principled
evaluation framework for algorithmic results comparison, and present
results of applying our implementation to real data.

@@

The numerator of the sum of squares ratio \(V\) quantifies the variation
between the two groups, while the denominator quantifies the variation
within each group. The sum of squares ratio \(V\) can be used as an
index for evaluating scans, metrics, and different sets of parameters if
the same data set is being used. Some examples will be presented in the
following section. If we impose the normality and independence
assumptions on the random effects \(e_{kl}\), the sum of squares ratio
\(V\) becomes a scaled F-statistic with degrees of freedom of \(k-1\)
and \(n_T - k\) and \(F = \frac{n_T - k}{k- 1} V\). If we want to
compare different data sets, stating that a certain setup can achieve
better separation on one data set than another, we can scale the sum of
squares ratio \(V\) and obtain the F-statistic to compare the p-values.

Using the sum of squares ratio \(V\) as an evaluation metric, we are
able to construct a pipeline that aims to find the optimal parameter
value for the CMPS algorithm by maximizing the sum of squares ratio. In
the following section, we will use this ratio to compare the CMPS
metrics introduced earlier and investigate effects of different sets of
parameters.

@@

As presented in the work of \citet{cmps}, researchers applied the CMPS
method to scans of one of the Hamby sets. While it is not explicitly
stated in the paper, we presume this to be Hamby 252, as only those
scans were publicly available at the time. In order to show that our
implementation of the CMPS algorithm is able to reproduce the results in
\citet{cmps} and be used for other data sets, we applied our
implementation to both Hamby set 252 and Hamby set 44. For the purpose
of reproducibility, we present how we obtained bullet signatures from
the Hamby set data and include code in the appendix: for both Hamby 252
and Hamby 44, we started with scans in the form of x3p files in the
database. Following the framework proposed by \citet{aoas}, we used the
same set of parameters, removed damaged bullet scans, obtained bullet
signatures for each bullet land engraving, and removed outliers in
bullet signatures. Note that researchers of \citet{cmps} applied the
CMPS algorithm to bullet signatures as well but used a framework
different from ours. However, since their work is not open-source, we
were not able to follow their framework and were only able to reproduce
the results for Hamby set 252 qualitatively.

@@

\hh{XXX Will, the conclusion needs to be a bit more than just a summary of what we did in the paper - think of it as a summary plus either the limitations of our approach, or next steps that should be done but goes beyond this paper.}

@@

\wj{Additionally, we propose new metrics based on the CMPS scores and compare the new metrics with the existing metrics. 
We also introduce a principled evaluation framework of algorithmic results using  a measure based on the sum of squares ratio. 
We showcase the implementation with two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare the CMPS algorithm using different sets of parameters and different metrics, following the evaluation framework based on the sum of squares ratio. }

The results obtained are promising: we were both able to reproduce the
results in \citet{cmps} qualitatively and achieve a clear separation
between the known match (KM) comparisons and known non-match (KNM)
comparisons in another bullet study.
\wj{However, the comparisons among different sets of parameters suggest that the optimal choice for parameter settings vary between different datasets. We propose an evaluation framework that allows a comprehensive cross-validation of parameter settings used in the algorithm for future research. 
Comparisons of the R implementation of the CMPS algorithm with the random forest model proposed by @aoas  suggest that adding the CMPS score as an additional feature in the random forest model might add further separation between known match (KM) comparisons and known non-match (KNM) comparisons.}

\wj{The open-source implementation of the CMPS algorithm provided by the \pkg{cmpsR} package is just one step towards in the framework of open science. Open science is particularly important to fields like forensic science where transparency and accuracy are critical to fair justice. 
Open-source implementations not only allow a peer-review but also facilitate further research, such as parameter cross-validations, method comparisons, and the development of statistical methods for modeling KM and KNM CMPS score distributions, which can then be used for error-rate estimations and fair applications that were called for by the PCAST [@pcast] report.
However, having open-source implementations is not enough for open science. 
In order to actually build the world of open science, many other efforts, such as open evaluation, open access publication, open educational resources, etc., are still needed. 
The database built and maintained by the National Institute of Standards and Technology is a good example of open data. 
What we are aiming for is an open system that is able to collect open results, compare multiple algorithms using multiple datasets, and evaluate algorithmic variation and accuracy. 
The evaluation metric we proposed in this paper can be used to compare different algorithms or even different datasets and is a snippet of this open system. 
But the core of this open system is the open science culture and contributions of the community.}

>>>>

According to the Uniform Crime Reporting Program of FBI (Federal Bureau
of Investigation), ``more than 76 percent (76.7) of the homicides for
which the FBI received weapons data in 2020
\hh{is this really 2020 or should it be 2019?}involved the use of
firearms'', and the number of cases that involved the use of firearms in
2020 increased by close to 30 percent (29.3) compared to that in 2019
\citep{fbiucr}. Firearm examination could play an important role in all
these cases. An important task of firearm examination is to answer the
question of whether two pieces of evidence come from the same source or
whether a piece of evidence matches a sample obtained from a specific
firearm. Here, in particular, we are interested to determine whether two
bullets were fired from the same gun barrel. Assessing the similarity
between two bullets is based on a comparison of striation marks acquired
during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of
similarity based on a visual comparison, generally, using a comparison
microscope \citep{afte}. This practice has been criticized for its lack
of objectivity and the associated problem of determining valid error
rates \citep{pcast}.

A report published by the \citet{nrc} states that ``{[}m{]}uch forensic
evidence-including, for example, bite marks and firearm and toolmark
identification---is introduced in criminal trials without any meaningful
scientific validation, determination of error rates, or reliability
testing to explain the limits of the discipline.'' To overcome those
criticisms and concerns, researchers have been making an effort to build
databases and develop frameworks and algorithms that bring an objective
and quantitative assessment into the field. The database of
\citet{nistdb} with digital 3d topographic scans from various studies by
\citet{brundage}, \citet{hamby}, \citet{Hamby:2019} and others, provides
a great resource for researchers. Algorithms that can be validated and
tested and generate quantitative results are developed for 2D and 3D
surface texture \citep{song2005}, tool marks
\citep{ChumbleyL_Scott2010VoTM}, and striae on Land Engraved Areas
(LEAs) of bullets \citep[\citet{pmid30444940}, \citet{cmps}]{aoas}. The
notion of bullet signatures is one of the results of these efforts, and
we will use it to explain the CMPS algorithm and showcase the
\pkg{cmpsR} implementation. Bullet signatures can be extracted from
bullet land engraved areas (LEAs), typically one for each land engraved
area, and how those bullet signatures are extracted from scans will be
discussed in the Background section. Bullet signatures capture striation
marks on the bullet in a numeric format and therefore serve as the
foundation for algorithms. These scans played an important role in
bringing objectivity into the field. In the following sections we will:
review the background of bullet signature comparisons, discuss how we
followed the idea described by \citet{cmps} for the implementation of
the CMPS algorithm, propose new metrics based on the CMPS score and a
principled evaluation framework for algorithmic results comparison, and
present results of applying our implementation to real data.

@@

The numerator of the sum of squares ratio \(V\) quantifies the variation
between the two groups, while the denominator quantifies the variation
within each group. The sum of squares ratio \(V\) can be used as an
index for evaluating scans, metrics, and different sets of parameters if
the same data set is being used. Some examples will be presented in the
following section. If we impose the normality and independence
assumptions on the random effects \(e_{kl}\), the sum of squares ratio
\(V\) becomes a scaled F-statistic with degrees of freedom of \(k-1\)
and \(n_T - k\) and \(F = \frac{n_T - k}{k- 1} V\). If we want to
compare different data sets, stating that a certain setup can achieve
better separation on one data set than another, we can scale the sum of
squares ratio \(V\) and obtain the F-statistic and obtain the
corresponding p-value as an index for comparison.

Using the sum of squares ratio \(V\) as an evaluation metric, we are
able to construct a pipeline that aims to find the optimal parameter
values for the CMPS algorithm by maximizing the sum of squares ratio.
Note that other measures of an algorithm, such as the accuracy and the
AUC (Area Under the Curve), are also important and useful. But when
algorithms achieve 100 percent accuracy and AUC value of 1, we need
other measures such as the sum of squares ratio to further distinguish
the performance of algorithms. In the following section, we will use the
sum of squares ratio to compare the CMPS metrics introduced earlier and
investigate effects of different parameter settings.

@@

As presented in the work of \citet{cmps}, researchers applied the CMPS
method to scans of one of the Hamby sets. While it is not explicitly
stated in the paper, we presume this to be Hamby 252, as only those
scans were publicly available at the time.
\hh{XXX Let's take out this statement once we hear back from NIST} In
order to show that our implementation of the CMPS algorithm is able to
reproduce the results in \citet{cmps} and be used for other data sets,
we applied our implementation to both Hamby set 252 and Hamby set 44.
For the purpose of reproducibility, we present how we obtained bullet
signatures from the Hamby set data and include code in the appendix: for
both Hamby 252 and Hamby 44, we started with scans in the form of x3p
files in the database. Following the framework proposed by \citet{aoas},
we used the same set of parameters, removed damaged bullet scans,
obtained bullet signatures for each bullet land engraving, and removed
outliers in bullet signatures. Note that researchers of \citet{cmps}
applied the CMPS algorithm to bullet signatures as well but used a
framework different from ours. However, since their work is not
open-source, we were not able to follow their framework and were only
able to reproduce the results for Hamby set 252 qualitatively.

@@

@@

Additionally, we propose two new metrics based on the CMPS scores and
compare the new metrics with the existing metrics. We also introduce a
principled evaluation framework of algorithmic results using a measure
based on the sum of squares ratio. We showcase the implementation with
two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare
the CMPS algorithm using different sets of parameters and different
metrics, following the evaluation framework based on the sum of squares
ratio. The results obtained are promising: we were both able to
reproduce the results in \citet{cmps} qualitatively and achieve a clear
separation between the known match (KM) comparisons and known non-match
(KNM) comparisons in another bullet study. However, the comparisons
among different sets of parameters suggest that the optimal choice for
parameter settings vary between different datasets.

Note, that the main difference between Hamby sets 44 and 252 is that
they were taken with different resolution scanning devices. The bullets
for Hamby 44 and Hamby 252 are fired from the same ten consecutively
rifled P-85 Ruger barrels. The difference in parameter settings for
optimal differentiation between same-source comparisons and
different-source comparisons is therefore quite surprising. As next
steps we need validation studies, similar to \citet{vanderplas} - trying
out the CMPS algorithm on firearms and ammunition combinations that are
different to test the limits of the algorithm. The evaluation framework
we proposed based on the sum of squares ratio will facilitate such
studies and other validation of the algorithm.

Comparisons of the R implementation of the CMPS algorithm with the
random forest model proposed by \citet{aoas} suggest that adding the
CMPS score as an additional feature in the random forest model might add
further separation between known match (KM) comparisons and known
non-match (KNM) comparisons.

The open-source implementation of the CMPS algorithm provided by the
\pkg{cmpsR} package is just one step towards in the framework of open
science. Open science is particularly important to fields like forensic
science where transparency and accuracy are critical to fair justice.
Open-source implementations not only allow a peer-review but also
facilitate further research, such as parameter cross-validations, method
comparisons, and the development of statistical methods for modeling KM
and KNM CMPS score distributions, which can then be used for error-rate
estimations and fair applications that were called for by the PCAST
\citep{pcast} report. However, having open-source implementations is not
enough for open science. In order to actually build the world of open
science, many other efforts, such as open evaluation, open access
publication, open educational resources, etc., are still needed. The
database built and maintained by the National Institute of Standards and
Technology is a good example of open data. What we are aiming for is an
open system that is able to collect open results, compare multiple
algorithms using multiple datasets, and evaluate algorithmic variation
and accuracy. The evaluation metric we proposed in this paper can be
used to compare different algorithms or even different datasets and is a
snippet of this open system. But the core of this open system is the
open science culture and contributions of the community.
